Analytical Predictions of Exoplanet Obliquities Generated by Planet-Disk Interactions

### 2020-02-24 writeups/Su_disk_disp.tex
Dong first pass

\begin{abstract}
    Large planetary spin-orbit misalignments (obliquities) are thought to play a
    key role in some planets' atmospheric circulation and tidal heating. A
    promising avenue to generate high obliquities is via spin-orbit resonances,
    where the spin and orbital precession frequencies of the planet become and
    remain commensurate as the system evolves. One such proposed mechanism
    involves a dissipating protoplanetary disk driving orbital precession of an
    interior planet \citep{millholland_disk}. This previous work is limited to
    small initial misalignment angles between the spin axis of the planet and
    orbital axis of the massive disk. In the present paper, this scenario is
    analytically characterized and extended to arbitrary misalignment angles. It
    is shown that (i) under adiabatic evolution, final planetary obliquities
    bifurcate into distinct tracks governed by interactions with the resonance,
    while (ii) under non-adiabatic evolution, a broad range of obliquities can
    be excited. Analytical formulae are provided for most interesting regimes of
    parameter space.
\end{abstract}

@@

Many mechanisms have been proposed to generate large planetary obliquities.
Giant impacts are traditionally used to generate the large obliquities of the
solar system ice giants \citep{original_gi, morbidelli_gi}. More recently,
obliquity excitations have been achieved via \emph{spin-orbit resonances}, where
the spin and orbital precession frequencies of the planet become and remain
commensurate as the system evolves (sometimes called \emph{resonance
advection}). Spin-orbit resonances have been studied extensively in solar system
dynamics as \emph{Cassini States} and are responsible for the significant
nonzero obliquities of both Jupiter ($3.1^\circ$) and Saturn ($26.7^\circ$)
\citep{colombo1966, henrard1987, ward2004I, ward_jupiter}. Recent work has
attempted to apply this mechanism to the ice giants as well
\citep{hamilton_tilting_ice}.

@@

However, their study was limited to small initial misalignment angles between
the spin vector of the planet and the orbital angular momentum of the
protoplanetary disk. Such an assumption is intuitive, since the planet forms
from the disk, but prospects for it to be violated exist. In particular, giant
and planetesimal impacts are expected to be abundant during planet formation and
may perturb the spin axis of the protoplanet from its primordial value
\citep{yalinewich2019atmospheric, schlichting2015atmospheric}. Furthermore, the
abundance of compact multi-planet systems provides support for possible
scattering between close-in, newly-formed planets \citep{usp_compact1,
usp_review}. It is conceivable that such processes could generate significant
spin-disk misalignment within the lifetime of the disk.

@@

The paper is organized as follows. In Section~\ref{s:eq}, we review relevant
spin-orbit dynamics and present the equations studied in the rest of the paper.
In Sections~\ref{s:ad} and~\ref{s:nonad}, we study the evolution of the system
when the disk either dissipates adiabatically or non-adiabatically respectively.
Exhaustive yet simple analytical arguments are able to explain completely the
numerical results in both limits. We discuss potential consequences in
Section~\ref{s:disc}. Our primary physical results consist of
Fig.~\ref{fig:ad_ensemble} in the adiabatic limit and
Fig.~\ref{fig:3_ensemble_05_15} in the non-adiabatic limit. Longer mathematical
calculations are relegated to the appendicies, including a leading-order
estimate of final planetary obliquities given small initial spin-disk
misalignment angles in Appendix~\ref{s:ad_approx}.

@@

We consider a star of mass $M_\star$ hosting an oblate planet of mass $M_{\rm
p}$, radius $R_{\rm p}$, at semimajor axis $a_{\rm p}$, and with spin angular
frequency $\Omega_{\rm p}$, and a protoplanetary disk of mass $M_{\rm d}$. For
simplicity, we treat the disk as a ring of radius $r_d$, but it is simple to
generalize to a disk with finite extent \citep[see][]{millholland_disk}.
Denote $\bm{S}$ the spin angular momentum and $\bm{L}$ the orbital angular
momentum of the planet, and $\bm{L}_{\rm d}$ the angular momentum of the disk.
The corresponding unit vectors are $\uv{s} \equiv \bm{S} / S$,
$\uv{l} \equiv \bm{L} / L$, and $\uv{l}_{\rm d} \equiv \bm{L}_{\rm
d} / L_d$.

We assume $S \ll L \ll L_{\rm d}$, so $\uv{l}_{\rm d}$ is approximately
constant and $\uv{l}$ experiences negligible backreaction torque from
$\uv{s}$. The equations of motion for $\uv{s}$ and $\uv{l}$
are
\begin{align}
    \rd{\uv{s}}{t} &= \omega_{\rm sl} \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        \equiv \alpha \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}},\label{eq:dsdt}\\
    \rd{\uv{l}}{t} &= \omega_{\rm ld}\p{\uv{l} \cdot \uv{l}_{\rm d}}
            \p{\uv{l} \times \uv{l}_{\rm d}}
        \equiv -g\p{\uv{l} \times \uv{l}_{\rm d}},
            \label{eq:dldt}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &= \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm
        p}} \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &= \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
        p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
% = 1/(152600 yr) (kp = 6kqp, Mstar = msun, mp = mJ, Rp = RJ, ap = AU, Wp = 2pi
% / (10 days))
% = 1 / (573362 yr) (Md/Mstar = 0.01, ap/rd = 1/30)
In Eq.~\eqref{eq:wsl}, $k_{\rm qp}$ and $k_{\rm p}$ are related to the planet's
rotation-induced quadrupole moment and moment of inertia respectively
\citep[see][]{lai2018}, and $n \equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the
planet's orbital mean motion. The three angles formed by the three vectors are
denoted
\begin{align}
    \theta &\equiv \arccos \p{\uv{s} \cdot \uv{l}},\\
    \theta_{\rm sd} &\equiv \arccos \p{\uv{s} \cdot \uv{l}_{\rm d}},\\
    I &\equiv \arccos \p{\uv{l} \cdot \uv{l}_{\rm d}}.
\end{align}
Following standard notation \citep[e.g.][]{colombo1966,peale1969,ward_jupiter},
we have defined $\alpha \equiv \omega_{\rm sl}$ and $g \equiv -\omega_{\rm
ld}\cos I$.

@@

Spin states satisfying $\p{\rd{\uv{s}}{t}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs). They require that $\uv{s}$, $\uv{l}$, and
$\uv{l}_{\rm d}$ be coplanar. There are either two or four CSs, depending
on the values of $\eta$. They are specified by the obliquity $\theta$ and the
precessional phase of $\uv{s}$ around $\uv{l}$, denoted by $\phi$. Following
the standard convention and nomenclature (see Figs.~\ref{fig:cs_vecs}
and~\ref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0$ and $\theta < 0$,
corresponding to $\uv{s}$ and $\uv{l}_{\rm d}$ being on opposite sides of
$\uv{l}$, while CS2 has $\phi = \pi$ and $\theta > 0$, corresponding to $\uv{s}$
and $\uv{l}_{\rm d}$ being on the same side of $\uv{l}$. When $\eta < \eta_{\rm
c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.
\begin{figure}
    \includegraphics[width=0.45\textwidth]{plots_diskdisp/2_3vec.png}
    \caption{Definition of angles in the Cassini state configuration and the
    adopted sign convention for $\theta$. Traditionally, $\theta \in [-\pi,
    pi]$.}\label{fig:cs_vecs}
\end{figure}
\begin{figure}
    %
    \includegraphics[width=0.45\textwidth]{plots_diskdisp/2_cs_locs.png}
    \caption{Cassini state obliquities as a function of $\eta$ for $I =
    5^\circ$. The vertical dashed line indicates $\eta_{\rm c}$ ($= 0.766$ for
    $I = 5^\circ$), where CS1 and CS4 merge and annihilate, and the horizontal
    dashed lines indicate $\theta = I$ and $I - 180^\circ$, the asymptotic
    values for CSs 2 and 3 for $\eta \gg \eta_{\rm c}$.}\label{fig:cs_locs}
\end{figure}

@@

The Hamiltonian (in the rotating frame) of the system is
\begin{align}
    \mathcal{H} &= -\frac{1}{2}\p{\uv{s} \cdot \uv{l}}^2
            + \eta \p{\uv{s} \cdot \uv{l}_{\rm d}}\nonumber\\
        &= -\frac{1}{2}\cos^2\theta
            + \eta \p{\cos \theta \cos I - \sin I \sin \theta \cos \phi}
                \label{eq:H}.
\end{align}
Trajectories in the phase space $\p{\phi, \cos \theta}$ satisfy $H = $ constant
(see Fig.~\ref{fig:eq_1contours}).

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two
trajectories originating and ending at CS4 are the only two infinite-period
orbits in the phase space. Together, these two critical trajectories are
referred to as the \emph{separatrix} and divide phase space into three zones. In
Fig.~\ref{fig:eq_1contours}, we show the separatrix, the three zones, and their
relations to the CSs. Note that trajectories in zone II librate about CS2
while those in zones I and III circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{plots_diskdisp/1contours_flip.png}
    \caption{Level curves plot of $\mathcal{H}\p{\phi, \cos \theta}$
    [Eq.~\eqref{eq:H}] for $I = 5^\circ$, where warmer colors denote more
    positive values. The black solid line is the separatrix, which only exists
    for $\eta < \eta_{\rm c} = 0.766$. The three zones (I, II, III), divided by
    the separatrix, are labeled. The Cassini states are denoted by filled
    circles and have the same colors as in Fig.~\ref{fig:cs_locs}. The interior
    of the separatrix, shaded in grey, is formally only defined for $\eta <
    \eta_{\rm c}$, but we may identify the points in phase space that flow into
    zone II when evolved forward in time (decreasing $\eta$ adiabatically); this
    is the shaded region in the top left panel, bounded by the black dotted
    line.}\label{fig:eq_1contours}
\end{figure*}

Since $\p{\phi, \cos \theta}$ form a set of canonical variables, the integral
$\oint \cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant
(see Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically
\citep{henrard1987,ward2004I}. Define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1}.
    \end{align}
\end{subequations}
The areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2}, \label{eq:A1}\\
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},
            \label{eq:A2}\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_{\rm c}$ since the
separatrix disappears, but a natural extension exists: evolve an initial phase
space point $p$ under adiabatic decrease of $\eta$ until the separatrix appears
at $\eta = \eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm
c}$. Since phase space area is conserved under adiabatic evolution, this
extension implies $A_{\rm i}\p{\eta > \eta_{\rm c}} = A_{\rm i}(\eta_{\rm c})$.
The boundary between these extended zones is denoted by the dashed black line in
the top left panel of Fig.~\ref{fig:eq_1contours}, where otherwise no separatrix
exists.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/1_areas.png}
    \caption{The solid lines show the fractional areas of each of the zones
    $A_{\rm i}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}. The colored
    dashed lines correspond to small $\eta$ approximations used in
    Appendix~\ref{s:ad_approx}. The colored dashed lines for $\eta > \eta_{\rm
    c}$ are the effective values of $A_{\rm II}, A_{\rm III}$ for $\eta >
    \eta_{\rm c}$, denoting the points that would flow into either area under
    adiabatic decrease of $\eta$ from $\eta > \eta_{\rm c}$. The vertical black
    dashed lines correspond to $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}] and
    the values of $\eta$ for which $A_{\rm II}$ is maximized ($\eta_{\rm \max,
    II}$) and for which $A_{\rm III}$ is minimized ($\eta_{\rm \min, III}$,
    Eq.~\eqref{eq:eta_minIII}). The horizontal black dashed line illustrates the
    fraction of phase space occupied by zone II at separatrix appearance.
    }\label{fig:eq_areas}
\end{figure}

@@

In this section, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ (or the disk mass $M_{\rm d}$; see Eqs.~\eqref{eq:eta}
and~\eqref{eq:deta_dt}) decreases sufficiently slowly that the evolution is
adiabatic. Strictly speaking, this requires the disk evolution timescale $t_{\rm
d}$ [Eq.~\eqref{eq:dmd_dt}] be much larger than all timesccales of the dynamical
system governed by the Hamiltonian [Eq.~\eqref{eq:H}]. This is of course not
possible in all cases, as the motion along the separatrix has an infinite
period.

@@

As shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it is the slowest characteristic frequency in the
system. The weak adiabaticity criterion can be recast as requiring that $\eta$
vary more slowly than this libration frequency
\citep[see][]{ward2004II,millholland_disk}, i.e.\ $t_{\rm d}^{-1} \lesssim
\omega_{\rm lib} / 2\pi$. In terms of the dimensionless parameter $\epsilon$
[Eq.~\eqref{eq:eps_def}], this adiabaticity condition is
\begin{equation}
    \epsilon = \frac{1}{\alpha t_{\rm d}}
        \lesssim \frac{1}{2\pi}\sqrt{\eta\sin I \sin \theta_2
            \p{1 + \eta \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{equation}
where $\theta_2$ is the obliquity at CS2. This formula differs from that given
in \citet{millholland_disk}, where the $\csc^3\theta_2$ term is neglected and
the square root is missing.

For $I = 5^\circ$, $\eta \sim \eta_{\rm c}$ Eq.~\eqref{eq:ad_constr} gives $\epsilon
\lesssim 0.085$. Since our criterion is only a weak condition for diabaticity,
we use $\epsilon = 3 \times 10^{-4}$ in our ``adiabatic'' simulations below
(Section~\ref{ss:ad_ensemble}). We explore the consequences of nonadiabatic
evolution in Section~\ref{s:nonad}.

@@

We consider the evolution of a system with arbitrary initial spin-disk
misalignment angle $\theta_{\rm sd, i}$ and initial $\eta_{\rm i} \gg 1$. We are
interested in the final spin obliquities $\theta_{\rm f}$ after $\eta$ gradually
decreases to $\eta_{\rm f} \ll 1$ (i.e.\ after the disk has dissipated to a
negligible mass). Note that when $\eta_{\rm i} \gg 1$, $\uv{l}$ precesses
around $\uv{l}_{\rm d}$ much faster than the spin-orbit precession
($\abs{\omega_{\rm ld}} \gg \abs{\omega_{\rm sl}}$), and the spin obliquity
$\theta$ varies rapidly. So it is more appropriate to use $\theta_{\rm sd}$
rather than $\theta$ to specify the initial spin orientation. We explore the
entire range $\theta_{\rm sd, i} \in [0, \pi]$ and choose $\epsilon = 3 \times
10^{-4}$ so that the system evolves almost adiabatically (see above).

>>>>

\begin{abstract}
    Large planetary spin-orbit misalignments (obliquities) are thought to play a
    key role in some planets' atmospheric circulation and tidal heating. A
    promising avenue to generate high obliquities is via spin-orbit resonances,
    where the spin and orbital precession frequencies of the planet become and
    remain commensurate as the system evolves. One such proposed mechanism
    involves a dissipating protoplanetary disk driving orbital precession of an
    interior planet \citep{millholland_disk}. This previous work is limited to
    small initial misalignment angles between the spin axis of the planet and
    orbital axis of the massive disk. In the present paper, this scenario is
    analytically characterized and extended to arbitrary misalignment angles. It
    is shown that (i) under adiabatic evolution, final planetary obliquities
    bifurcate into distinct tracks governed by interactions with the resonance,
    while (ii) under nonadiabatic evolution, a broad range of obliquities can
    be excited. Analytical formulae are provided for most interesting regimes of
    parameter space.
\end{abstract}

@@

Many mechanisms have been proposed to generate large planetary obliquities.
Giant impacts are traditionally used to generate the large obliquities of the
solar system ice giants \citep{original_gi, morbidelli_gi}. More recently,
obliquity excitations have been achieved via \emph{spin-orbit resonances}, where
the spin and orbital precession frequencies of the planet become and remain
commensurate as the system evolves (sometimes called \emph{resonance
advection}). Spin-orbit resonances have been studied extensively in solar system
dynamics as \emph{Cassini States} and are thought to be responsible for the
significant nonzero obliquities of both Jupiter ($3.1^\circ$) and Saturn
($26.7^\circ$) \citep{colombo1966, henrard1987, ward2004I, ward_jupiter}. Recent
work has attempted to apply this mechanism to the ice giants as well
\citep{hamilton_tilting_ice}.

@@

However, their study was limited to small initial misalignment angles between
the spin vector of the planet and the orbital angular momentum of the
protoplanetary disk. Such an assumption is intuitive, since the planet forms
from the disk, but evidence suggests it may not always be satisfied. In
particular, giant and planetesimal impacts are expected to be abundant during
planet formation and may perturb the spin axis of the protoplanet from its
primordial value \citep{yalinewich2019atmospheric, schlichting2015atmospheric}.
Furthermore, the prevalence of compact multi-planet systems provides support for
possible scattering between close-in, newly-formed planets \citep{usp_compact1,
usp_review}. It is conceivable that such processes could generate significant
spin-disk misalignment within the lifetime of the disk.

@@

The paper is organized as follows. In Section~\ref{s:eq}, we review relevant
spin-orbit dynamics and present the equations studied in the rest of the paper.
In Sections~\ref{s:ad} and~\ref{s:nonad}, we study the evolution of the system
when the disk either dissipates adiabatically or nonadiabatically respectively.
Exhaustive yet simple analytical arguments are able to explain completely the
numerical results in both limits. We discuss potential consequences in
Section~\ref{s:disc}. Our primary physical results consist of
Fig.~\ref{fig:ad_ensemble} in the adiabatic limit and
Fig.~\ref{fig:3_ensemble_05_15} in the nonadiabatic limit. Longer mathematical
calculations are relegated to the appendicies, including a leading-order
estimate of final planetary obliquities given small initial spin-disk
misalignment angles in Appendix~\ref{s:ad_approx}.

@@

We consider a star of mass $M_\star$ hosting an oblate planet of mass $M_{\rm
p}$, radius $R_{\rm p}$, at semimajor axis $a_{\rm p}$, and with spin angular
frequency $\Omega_{\rm p}$, and a protoplanetary disk of mass $M_{\rm d}$. For
simplicity, we treat the disk as a ring of radius $r_{\rm d}$, but it is simple to
generalize to a disk with finite extent \citep[see][]{millholland_disk}.
Denote $\bm{S}$ the spin angular momentum and $\bm{L}$ the orbital angular
momentum of the planet, and $\bm{L}_{\rm d}$ the angular momentum of the disk.
The corresponding unit vectors are $\uv{s} \equiv \bm{S} / S$,
$\uv{l} \equiv \bm{L} / L$, and $\uv{l}_{\rm d} \equiv \bm{L}_{\rm
d} / L_{\rm d}$.

We assume $S \ll L \ll L_{\rm d}$, so $\uv{l}_{\rm d}$ is approximately
constant and $\uv{l}$ experiences negligible backreaction torque from
$\uv{s}$. The equations of motion for $\uv{s}$ and $\uv{l}$
are
\begin{align}
    \rd{\uv{s}}{t} &= \omega_{\rm sl} \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        \equiv \alpha \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}},\label{eq:dsdt}\\
    \rd{\uv{l}}{t} &= \omega_{\rm ld}\p{\uv{l} \cdot \uv{l}_{\rm d}}
            \p{\uv{l} \times \uv{l}_{\rm d}}
        \equiv -g\p{\uv{l} \times \uv{l}_{\rm d}},
            \label{eq:dldt}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &= \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm
        p}} \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &= \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
        p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
% = 1/(152600 yr) (kp = 6kqp, Mstar = msun, mp = mJ, Rp = RJ, ap = AU, Wp = 2pi
% / (10 days))
% = 1 / (573362 yr) (Md/Mstar = 0.01, ap/rd = 1/30)
In Eq.~\eqref{eq:wsl}, $k_{\rm qp}$ and $k_{\rm p}$ are related to the planet's
rotation-induced quadrupole moment and moment of inertia respectively
\citep[see][]{lai2018}, and $n \equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the
planet's orbital mean motion. The three angles formed by the three vectors are
denoted
\begin{align}
    \theta &\equiv \arccos \p{\uv{s} \cdot \uv{l}},\\
    \theta_{\rm sd} &\equiv \arccos \p{\uv{s} \cdot \uv{l}_{\rm d}},\\
    I &\equiv \arccos \p{\uv{l} \cdot \uv{l}_{\rm d}}.
\end{align}
Following standard notation \citep[e.g.][]{colombo1966,peale1969,ward2004I}, we
have defined $\alpha \equiv \omega_{\rm sl}$ and $g \equiv -\omega_{\rm ld}\cos
I$.

@@

Spin states satisfying $\p{\rd{\uv{s}}{\tau}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs). They require that $\uv{s}$, $\uv{l}$, and
$\uv{l}_{\rm d}$ be coplanar. There are either two or four CSs, depending
on the value of $\eta$. They are specified by the obliquity $\theta$ and the
precessional phase of $\uv{s}$ around $\uv{l}$, denoted by $\phi$. Following
the standard convention and nomenclature (see Figs.~\ref{fig:cs_vecs}
and~\ref{fig:cs_locs}), CSs 1, 3, 4 have $\phi = 0$ and $\theta < 0$,
corresponding to $\uv{s}$ and $\uv{l}_{\rm d}$ being on opposite sides of
$\uv{l}$, while CS2 has $\phi = \pi$ and $\theta > 0$, corresponding to $\uv{s}$
and $\uv{l}_{\rm d}$ being on the same side of $\uv{l}$. When $\eta < \eta_{\rm
c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.
\begin{figure}
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/2_3vec.png}
    \caption{Definition of angles in the Cassini state configuration and the
    adopted sign convention for $\theta$. Traditionally, $\theta \in [-\pi,
    \pi]$.}\label{fig:cs_vecs}
\end{figure}
\begin{figure}
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/2_cs_locs.png}
    \caption{Cassini state obliquities as a function of $\eta$ for $I =
    5^\circ$. The vertical dashed line indicates $\eta_{\rm c}$ ($= 0.766$ for
    $I = 5^\circ$), where CS1 and CS4 merge and annihilate, and the horizontal
    dashed lines indicate $\theta = I$ and $I - 180^\circ$, the asymptotic
    values for CSs 2 and 3 for $\eta \gg \eta_{\rm c}$.}\label{fig:cs_locs}
\end{figure}

@@

The Hamiltonian (in the rotating frame) of the system is
\begin{align}
    \mathcal{H}\p{\phi, \cos \theta} &= -\frac{1}{2}\p{\uv{s} \cdot \uv{l}}^2
            + \eta \p{\uv{s} \cdot \uv{l}_{\rm d}}\nonumber\\
        &= -\frac{1}{2}\cos^2\theta
            + \eta \p{\cos \theta \cos I - \sin I \sin \theta \cos \phi}
                \label{eq:H}.
\end{align}
Here, $\phi$ and $\cos \theta$ are canonically conjugate variables. Trajectories
in the phase space $\p{\phi, \cos \theta}$ satisfy $H = $ constant (see
Fig.~\ref{fig:eq_1contours}).

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two
trajectories originating and ending at CS4 are the only two infinite-period
orbits in the phase space. Together, these two critical trajectories are
referred to as the \emph{separatrix} and divide phase space into three zones. In
Fig.~\ref{fig:eq_1contours}, we show the separatrix, the three zones, and their
relations to the CSs. Trajectories in zone II librate about CS2 while those in
zones I and III circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{plots_diskdisp/1contours_flip.png}
    \caption{Level curves plot of $\mathcal{H}\p{\phi, \cos \theta}$
    [Eq.~\eqref{eq:H}] for $I = 5^\circ$, where warmer colors denote more
    positive values. The black solid line is the separatrix, which only exists
    for $\eta < \eta_{\rm c} = 0.766$. The three zones (I, II, III), divided by
    the separatrix, are labeled. The Cassini states are denoted by filled
    circles and have the same colors as in Fig.~\ref{fig:cs_locs}. The interior
    of the separatrix, shaded in grey, is formally only defined for $\eta <
    \eta_{\rm c}$, but we may identify the points in phase space that flow into
    zone II when evolved forward in time (decreasing $\eta$ adiabatically); this
    is the shaded region in panel (a), bounded by the black dotted
    line.}\label{fig:eq_1contours}
\end{figure*}

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically
\citep{henrard1987,ward2004I}. Define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1}.
    \end{align}
\end{subequations}
The areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2}, \label{eq:A1}\\
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},
            \label{eq:A2}\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$A_{\rm i}\p{\eta > \eta_{\rm c}} = A_{\rm i}(\eta_{\rm c})$. The boundary
between these extended zones is denoted by the dashed black line in panel (a) of
Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{plots_diskdisp/1_areas.png}
    \caption{The solid lines show the fractional areas of each of the zones
    $A_{\rm i}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}. The colored
    dashed lines correspond to small $\eta$ approximations used in
    Appendix~\ref{s:ad_approx}. The colored dashed lines for $\eta > \eta_{\rm
    c}$ are the effective values of $A_{\rm II}, A_{\rm III}$ for $\eta >
    \eta_{\rm c}$, denoting the points that would flow into either area under
    adiabatic decrease of $\eta$ from $\eta > \eta_{\rm c}$. The vertical black
    dashed lines correspond to $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}] and
    the values of $\eta$ for which $A_{\rm II}$ is maximized ($\eta_{\rm \max,
    II}$) and for which $A_{\rm III}$ is minimized ($\eta_{\rm \min, III}$,
    Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

In this section, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ [or the disk mass $M_{\rm d}$; see Eqs.~\eqref{eq:eta}
and~\eqref{eq:deta_dt}] decreases sufficiently slowly that the evolution is
adiabatic. Strictly speaking, this requires the disk evolution timescale $t_{\rm
d}$ [Eq.~\eqref{eq:dmd_dt}] be much larger than all timesccales of the dynamical
system governed by the Hamiltonian [Eq.~\eqref{eq:H}]. This is of course not
possible in all cases, as the motion along the separatrix has an infinite
period.

@@

As shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it is the slowest characteristic frequency in the
system. The weak adiabaticity criterion can be recast as requiring that $\eta$
vary more slowly than this libration frequency
\citep[see][]{ward2004II,millholland_disk}, i.e.\ $t_{\rm d}^{-1} \lesssim
\omega_{\rm lib} / 2\pi$. In terms of the dimensionless parameter $\epsilon$
[Eq.~\eqref{eq:eps_def}], this adiabaticity condition is
\begin{equation}
    \epsilon = \frac{1}{\alpha t_{\rm d}}
        \lesssim \frac{1}{2\pi}\sqrt{\eta\sin I \sin \theta_2
            \p{1 + \eta \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{equation}
where $\theta_2$ is the obliquity at CS2. This formula differs from that given
in \citet{millholland_disk}, where the $\csc^3\theta_2$ term is neglected and
the square root is missing\footnote{The missing $\csc^3 \theta_2$ term can be
traced to a $\theta \gg I$ approximation made in Eq.~(3) of \citet{ward2004II}.
Since $\theta_2 \approx I$ for $\eta \gg 1$ (Fig.~\ref{fig:cs_locs}), this
approximation is not always valid.}.

For $I = 5^\circ$, $\eta \sim \eta_{\rm c}$ Eq.~\eqref{eq:ad_constr} gives
$\epsilon \lesssim 0.085$. Since our criterion is only a weak condition for
diabaticity, we use $\epsilon = 3 \times 10^{-4}$ in our ``adiabatic''
simulations below. We explore the consequences of nonadiabatic evolution in
Section~\ref{s:nonad}.

@@

We consider the evolution of a system with arbitrary initial spin-disk
misalignment angle $\theta_{\rm sd, i}$ and initial $\eta_{\rm i} \gg 1$. We are
interested in the final spin obliquities $\theta_{\rm f}$ after $\eta$ gradually
decreases to $\eta_{\rm f} \ll 1$ (i.e.\ after the disk has dissipated to a
negligible mass). Note that when $\eta_{\rm i} \gg 1$, $\uv{l}$ precesses
around $\uv{l}_{\rm d}$ much faster than the spin-orbit precession
($\abs{\omega_{\rm ld}} \gg \abs{\omega_{\rm sl}}$), and the spin obliquity
$\theta$ varies rapidly. It is thus more appropriate to use $\theta_{\rm sd, i}$
rather than $\theta$ to specify the initial spin orientation. We explore the
entire range $\theta_{\rm sd, i} \in [0, \pi]$ and choose $\epsilon = 3 \times
10^{-4}$ so that the system evolves almost adiabatically (see above).
